---
title: "Rain Beetles v. Missoula Floods"
author: "[John Powers](http://sites.uci.edu/powers/)"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
    toc: yes
    toc_float: TRUE
editor_options: 
  chunk_output_type: console
---

# Purpose

Map *Pleocoma* distributions and evaluate whether the Pleistocene Missoula floods impacted their distribution in the Willamette Valley.

```{r setup, message=FALSE}
library(tidyverse)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, fig.path = "images/")

library(sf)
mycrs <- "+proj=longlat +datum=WGS84"

beetle_taxa <- c("Coleoptera", "Pleocomidae", "Scarabaeidae", "Lucanidae") 
library(RColorBrewer)
beetle_pal <- set_names(brewer.pal(4, "Set1"), beetle_taxa)
```

# Missoula Floods highpoint

Multiple sources place the highest extent of the \~40 Missoula Floods in the Willamette Valley at an elevation of 122 m, although this was locally higher in the Portland Basin as the 150 m Rocky Butte saddle was overtopped (references in [Minervini et al. 2003](https://doi.org/10.3133/ofr2003408)).

This flood extent has been mapped against LIDAR data in the Portland Basin ([Burns and Coe 2012](https://pubs.oregon.gov/dogami/ims/p-ims-036.htm)).

```{r flood}
maxflood <-  122 #in meters
res <- 6 #resolution of elevation histograms = maxflood / res

flood <- data.frame(x = c(-124, -122), y = c(43.5, 46)) #box around the Willamette Valley
flood.sf <- flood %>% st_as_sf(coords = c("x", "y"), crs = mycrs) %>% st_bbox() %>% st_as_sfc()
```

# Beetle records from GBIF

Search GBIF for:

1.  rain beetles across their range: [GBIF download 2023-11-10](https://doi.org/10.15468/dl.kd2swv)
2.  all beetles, and two other scarab families around the Willamette Valley for sampling effort comparison

```{r gbif_search, eval=FALSE}
library(rgbif)

#GBIF download in simple format, used for annotation: data/0055359-231002084531237.csv
Pleocomidae.gbif <- occ_search(taxonKey = name_backbone("Pleocomidae")$usageKey, limit=1e4)$data %>% 
  write_csv("data/Pleocomidae_gbif.csv")

Coleoptera.gbif.flood <- map_dfr(set_names(beetle_taxa)[-2], .id="taxon",
                                 ~ occ_search(taxonKey = name_backbone(.x)$usageKey, geometry=st_as_text(flood.sf), limit=1e4)$data) %>%  
  write_csv("data/Coleoptera_gbif_flood.csv.gz")
```

Add hand-annotated localities, coordinates, and uncertainty estimates for rain beetle occurrences.

```{r gbif_annotate}
#Sheet with JP's hand-annotated localities, coordinates, and uncertainty estimates
Pleocomidae.gbif.newloc <- read_csv("data/Pleocoma log - GBIF20231110newloc.csv") %>%
  select(gbifID, newLocality,	newLocalityScale,	newCoordinateUncertaintyInMeters,	newLatLong) %>% 
  separate(newLatLong, into=c("newLat","newLong"), sep=", ", convert=T)

#Replace some GBIF data with hand annotations
Pleocomidae.gbif.fix <- read_csv("data/Pleocomidae_gbif.csv", guess_max = 2000, col_types=list(recordNumber=col_character())) %>% 
  left_join(Pleocomidae.gbif.newloc) %>% 
  mutate(locality=if_else(!is.na(newLocality), newLocality, locality),
         decimalLatitude=if_else(!is.na(newLat), newLat, decimalLatitude),
         decimalLongitude=if_else(!is.na(newLat), newLong, decimalLongitude),
         coordinateUncertaintyInMeters=if_else(!is.na(newCoordinateUncertaintyInMeters), 
                                               newCoordinateUncertaintyInMeters, coordinateUncertaintyInMeters), 
         taxon = "Pleocomidae", .keep="unused")

#filter out uncertain coordinates, including vague localities (>10 km uncertainty) and obscured iNat coordinates
Pleocomidae.gbif.goodcoords <- Pleocomidae.gbif.fix %>% 
  drop_na(decimalLongitude) %>% 
  filter(decimalLongitude < -100, #drop fossils in Asia
    is.na(coordinateUncertaintyInMeters) | coordinateUncertaintyInMeters <= 10000)

#Convert to spatial objects
Pleocomidae.gbif.sf <- Pleocomidae.gbif.goodcoords %>% 
  st_as_sf(coords=c("decimalLongitude","decimalLatitude"), crs=mycrs)
Pleocomidae.gbif.flood <- st_intersection(Pleocomidae.gbif.sf, flood.sf)

gbif.flood <- read_csv("data/Coleoptera_gbif_flood.csv.gz", guess_max=20000, col_types=list(georeferencedDate=col_character())) %>% 
  st_as_sf(coords=c("decimalLongitude","decimalLatitude"), crs=mycrs) %>%
  bind_rows(Pleocomidae.gbif.flood)
```

# Elevations

Get elevations of all occurrences and a raster elevation map of the flood area.

```{r elevation, eval=FALSE}
library(elevatr)
#zoom level sets resolution, see https://github.com/tilezen/joerd/blob/master/docs/data-sources.md#what-is-the-ground-resolution
Pleocomidae.gbif.sf %>% select(geometry, gbifID) %>% 
  get_elev_point(src = "aws", z=8) %>% as_tibble() %>% write_csv("data/Pleocomidae_gbif_elev.csv")
gbif.flood %>% select(geometry, gbifID) %>% 
  get_elev_point(src = "aws", z=11) %>% as_tibble() %>% write_csv("data/gbif_flood_elev.csv")
```

```{r raster, cache=FALSE}
library(terra)
library(tidyterra)
library(elevatr)
flood.raster <- get_elev_raster(gbif.flood, src = "aws", z=8) %>% rast() %>% crop(ext(unlist(flood)))
```

Export [table of rain beetle occurrences in the flood area](./data/Pleocomidae_gbif_flood.csv) for use in Google Earth.

Imported the CSV and followed [these steps](https://googleearthdesign.blogspot.com/2011/01/flood-simulation-howto.html) for flooding an area in Google Earth to produce this [KML file](./data/Pleocomidae_gbif_flood.kml).

```{r export}
Pleocomidae.gbif.sf <- Pleocomidae.gbif.sf %>% select(-elevation) %>% left_join(read_csv("data/Pleocomidae_gbif_elev.csv", col_select = -geometry))
gbif.flood <- gbif.flood %>% select(-elevation) %>% left_join(read_csv("data/gbif_flood_elev.csv", col_select = -geometry) %>% distinct(gbifID, .keep_all = T))

Pleocomidae.gbif.flood <- bind_cols(as_tibble(gbif.flood), st_coordinates(gbif.flood)) %>% 
  filter(taxon =="Pleocomidae") %>% 
  select(gbifID, family, species, infraspecificEpithet, sex, eventDate, 
         decimalLatitude=Y, decimalLongitude=X, coordinateUncertaintyInMeters, elevation,
         locality, newLocalityScale, institutionCode, references, occurrenceRemarks, identificationRemarks) %>% 
  write_csv("data/Pleocomidae_gbif_flood.csv")

library(knitr)
Pleocomidae.gbif.flood %>% filter(elevation <= maxflood, decimalLatitude>43.7) %>% arrange(desc(decimalLatitude)) %>% 
  kable(caption="Rain beetles found at or below the highest flood")
```

# Maps

```{r map_all, fig.width=10}
ggplot(Pleocomidae.gbif.sf) + 
  geom_polygon(data = map_data("state"), aes(x = long, y = lat, group = group)) +
  geom_sf(aes(color=species)) + coord_sf(xlim=c(-125, -115), ylim=c(30, 50)) + theme_void()

ggplot(Pleocomidae.gbif.sf, aes(x=elevation, fill=species)) + facet_wrap(vars(species), scales="free_y") + 
  geom_histogram(binwidth=100) + guides(fill="none")
```

```{r map_NW}
ggplot(Pleocomidae.gbif.sf %>% filter(stateProvince!="California")) + 
  geom_polygon(data = map_data("state"), aes(x = long, y = lat, group = group), fill = "grey95", color = "gray40") +
  geom_sf(aes(color=species)) + coord_sf(xlim=c(-125, -120), ylim=c(42.5, 47)) + theme_void() 
```

```{r map_flood, fig.height=12, fig.width=10}
land <- colorRampPalette(rev(brewer.pal(4, "Greys")))
flooded <- colorRampPalette(brewer.pal(4, "Blues"))

ggplot(gbif.flood) + 
  geom_spatraster(data=flood.raster)+ 
  geom_sf(aes(color=taxon, alpha=taxon=="Pleocomidae"), shape=19)+  
  theme_void() + guides(alpha="none", fill="none") + scale_color_manual("", values=beetle_pal)

ggplot(gbif.flood %>% filter(taxon=="Pleocomidae")) + 
  geom_spatraster(data=flood.raster)+ 
  scale_fill_gradientn(colors=c(flooded(maxflood-minmax(flood.raster)[1]), land(minmax(flood.raster)[2]-maxflood)))+
  scale_color_manual(values=c("black","white")) + scale_shape_manual(values=19)+
  geom_sf(aes(color=elevation>maxflood))+#, shape = if_else(elevation>maxflood, "1",NA))) + 
  #geom_sf_text(aes(color=elevation>maxflood, label=if_else(elevation<maxflood, elevation - maxflood,NA)))+
  theme_void() + guides(color="none",fill="none", shape="none")

flood.north <- data.frame(x = c(-123.5, -122.4), y = c(44.5,45.7))
flood.raster.north <- flood.raster %>% crop(ext(unlist(flood.north)))
ggplot(gbif.flood %>% filter(taxon=="Pleocomidae")) + 
  geom_spatraster(data=flood.raster.north) +
  scale_fill_gradientn(colors=c(flooded(maxflood-minmax(flood.raster.north)[1]), land(minmax(flood.raster.north)[2]-maxflood)))+
  scale_color_manual(values=c("black","white")) + scale_shape_manual(values=19)+
  geom_sf(aes(color=elevation>maxflood))+#, shape = if_else(elevation>maxflood, "1",NA))) + 
  #geom_sf_text(aes(color=elevation>maxflood, label=if_else(elevation<maxflood, elevation - maxflood,NA)))+
  theme_void() + guides(color="none",fill="none", shape="none") +
  coord_sf(ylim=flood.north$y, xlim=flood.north$x)
```

# Elevation distribution

Elevation distributions of beetle observations:

1.  Raw counts at each elevation
2.  Divided by land area at each elevation
3.  Divided by total beetle observations at each elevation

```{r sampling}
gbif.elev.compare <- bind_rows(bind_cols(as_tibble(gbif.flood), st_coordinates(gbif.flood)) %>% 
                                 distinct(taxon, elevation, X, Y), #takes out some collections that have multiple records for the same spot
                               tibble(taxon="area", elevation = sample(values(flood.raster), 2e4)))

ggplot(gbif.elev.compare, aes(x=elevation, fill=taxon)) + facet_wrap(vars(taxon), scales="free_y", ncol=1) + 
  geom_histogram(binwidth=maxflood/res, boundary=maxflood) +  geom_vline(xintercept=maxflood) + 
  guides(fill="none")  + scale_x_continuous(n.breaks = 13) + coord_cartesian(xlim=c(NA,maxflood*10)) + theme_minimal() +
  labs(x="Elevation (m)", y="Occurrences", title="Elevation distribution of beetle records and land in flood region") +
  scale_fill_manual(values=c(beetle_pal,area="black"))

gbif.elev.perarea <- gbif.elev.compare %>% 
  mutate(elev.binned = (cut(elevation, breaks = seq(0, maxflood*10, by=maxflood/res), labels=F)-1)*maxflood/res) %>% 
  count(taxon, elev.binned) %>% pivot_wider(names_from=taxon, values_from=n) %>% 
  mutate(across(-elev.binned, ~replace_na(.x/area,0))) %>% select(-area)

gbif.elev.perarea %>% pivot_longer(all_of(beetle_taxa), names_to = "taxon", values_to="obs_per_area") %>% 
  ggplot(aes(x=elev.binned, fill=taxon, y=obs_per_area)) + facet_wrap(vars(taxon), scales="free_y", ncol=1) + 
  geom_col(just=0)+ geom_vline(xintercept=maxflood) + 
  guides(fill="none") + scale_x_continuous(n.breaks = 13) + coord_cartesian(xlim=c(NA,maxflood*10)) + theme_minimal() +
  labs(x="Elevation (m)", y="Occurrences per area", title="Records of beetles per land area")+
  scale_fill_manual(values=beetle_pal)

gbif.elev.perarea %>% filter(Coleoptera>0) %>% 
  mutate(across(-elev.binned, ~.x/Coleoptera)) %>% select(-Coleoptera) %>% 
  pivot_longer(any_of(beetle_taxa), names_to = "taxon", values_to="obs_per_beetle") %>% 
  ggplot(aes(x=elev.binned, fill=taxon, y=obs_per_beetle)) + facet_wrap(vars(taxon), scales="free_y", ncol=1) + 
  geom_col(just=0)+ geom_vline(xintercept=maxflood) + 
  guides(fill="none") + scale_x_continuous(n.breaks = 13) + coord_cartesian(xlim=c(NA,maxflood*10)) + theme_minimal() +
  labs(x="Elevation (m)", y="Occurrences per beetle occurrence", title="Records of beetles relative to total records")+
  scale_fill_manual(values=beetle_pal)
```

# Seasonality

```{r seasons}
Pleocomidae.gbif.goodcoords %>% 
  mutate(region= cut(decimalLatitude, c(0,35,42,90), labels=c("SoCal","NorCal","Oregon")),
         month = month(eventDate, label=T))%>% drop_na(month) %>% 
  ggplot(aes(x=month, fill=region)) + facet_wrap(vars(region), ncol=1, scales="free_y") +
  geom_bar() + labs(x="Month", y="Occurrences", title="Rain beetle emergence by region") + theme_minimal() + guides(fill="none")
```
